---
title: "GVTP"
output:
  html_document:
    toc: true
    toc_depth: 3
    toc_float: true
    code_folding: hide
    self_contained: false
editor_options: 
  chunk_output_type: console
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = T, warning = F, message = F)
```

## Skim

```{r}
# load libraries ----
if (!require(librarian)){
  install.packages(librarian)
  library(librarian)
}
# load libraries, install if missing
shelf(
  # reporting
  DT, htmltools,
  # sdm
  corrplot, dismo, sdm,
  # tidyverse
  dplyr, readr, readxl, tidyr,
  # utilities
  fs, glue, here, skimr)
# sdm::installAll()
select = dplyr::select

# paths & variables ----
dir_gdrive <- case_when(
  Sys.info()[['user']] == "bbest" ~ "/Volumes/GoogleDrive/My Drive/projects/dudek")
d_xls    <- path(dir_gdrive, "data/gvtp_hab_char_data_compiled_Final.xls")
cols_xls <- path(dir_gdrive, "data/gvtp_hab_char_columns.xlsx")
# excel_sheets(d_xls)

# read sheets of data
plot   <- read_excel(d_xls, "Plot")
spp    <- read_excel(d_xls, "Species")
soil   <- read_excel(d_xls, "Soil Layers")
veg    <- read_excel(d_xls, "Veg")
herbht <- read_excel(d_xls, "Herb Height")
gvtpht <- read_excel(d_xls, "GVTP Height Heads")
rdm    <- read_excel(d_xls, "RDM")

# skim sheets
h3("plot")
skim(plot)
h3("spp")
skim(spp)
h3("soil")
skim(soil)
h3("veg")
skim(veg)
h3("herbht")
skim(herbht)
h3("gvtpht")
skim(gvtpht)
h3("rdm")
skim(rdm)
```

## SDM

Species distribution modeling...

### Prep predictors

```{r}
d <- plot %>% 
  filter(!is.na(gvtp_count)) %>% 
  mutate(
    gvtp = as.integer(!gvtp_count == "0")) %>% 
  select_if(is.numeric)

predictors <- setdiff(names(d), "gvtp") %>% sort()

d <- d %>% 
  select(gvtp, all_of(predictors))

# abbreviate names
d_cols_abbr <- abbreviate(names(d), minlength=4)
names(d) <- d_cols_abbr
d_cols <- tibble(
  column_original = names(d_cols_abbr),
  column_abbreviated = d_cols_abbr)
write_csv(d_cols, cols_xls)
datatable(d_cols)

datatable(d)
```

### Correlation among predictors

* [Introduction to SDMs: data preparation and simple model fitting](https://damariszurell.github.io/HU-GCIB/3_SDM_intro.html)

```{r}
# We first estimate a correlation matrix from the predictors. We use Spearman rank correlation coefficient, as we do not know whether all variables are normally distributed.
cor_mat <- cor(d[,-1], method='spearman')

# We can visualise this correlation matrix. For better visibility, we plot the correlation coefficients as percentages.
corrplot.mixed(cor_mat, tl.pos='lt', tl.cex=0.5, number.cex=0.4, addCoefasPercent=T)
```

### Fit models

* [sdm](http://www.biogeoinformatics.org/)

```{r}
d_sdm <- sdmData(gvtp ~ ., train = as.data.frame(d))
d_sdm

m1 <- sdm(
  gvtp ~ ., data = d_sdm, 
  methods = c("glm", "gam", "brt"))
m1

m2 <- sdm(
  gvtp ~ .,data = d_sdm,
  methods = c("rf", "tree", "fda", "mars", "svm"),
  replication = "sub", test.percent = 30, n = 2)

m2
getModelInfo(m2)

roc(m2)

roc(m2,smooth=T)
```

### Importance of predictors

```{r}
# gui(m2)

vi <- getVarImp(m2, id = 1, wtest = "training")

plot(vi,'cor')

vi@varImportance %>% 
  arrange(desc(corTest)) %>% 
  datatable()
```

### gam

```{r}
group      <- kfold(d, 5)
pres_train <- d[group != 1, ]
pres_test  <- d[group == 1, ]

maxent()
## Loading required namespace: rJava
## This is MaxEnt version 3.4.1
xm <- maxent(predictors, pres_train, factors='biome')
plot(xm)

```


## Next Steps

- subset of variables based on 1 per group in [gvtp_hab_char_columns.xlsx - Google Sheets](https://docs.google.com/spreadsheets/d/1LUjoRMollGuouUyfQeoAPbYC6Hl6i7kT/edit#gid=1414489956)
- want term plots:
  - gam
  - [R: Maxent](http://finzi.psych.upenn.edu/R/library/dismo/html/maxent.html)
- [Species distribution modeling — R Spatial](https://rspatial.org/raster/sdm/index.html)
- [6 Available Models | The caret Package](http://topepo.github.io/caret/available-models.html)
- [Lab 6. Species](http://benbestphd.com/landscape-ecology-labs/lab6.html)

## References

[Zotero group: dudek-gvtp](https://www.zotero.org/groups/2587262/dudek-gvtp/library)

