---
title: ""
output:
  officedown::rdocx_document: default
  word_document: default
  html_document: default
editor_options: 
  chunk_output_type: console
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(warning = F, echo = F, message = F, eval = F)
```

```{r, eval = T}
source("setup.R") # d, d_cols
shelf(
  flextable, cowplot, officedown, officer)
```

# Variable Selection

* [Introduction to SDMs: data preparation and simple model fitting](https://damariszurell.github.io/HU-GCIB/3_SDM_intro.html#41_fitting_our_first_glm)

```{r, eval = T}
mdls_all_lnd <- get_mdls(resp = "g_pres", cat = "landscape")

cor_lnd <- get_cor(d, mdls_all_lnd$pred)

mdls_lnd <- filt_mdls_cor(mdls_all_lnd, cor_lnd)

# check that no correlations still exist
cor2_lnd <- get_cor(d, mdls_lnd$pred)
stopifnot(nrow(cor2_lnd) == 0)
```

### Maxent

* [Maxent -- Modeling methods | R Spatial](https://rspatial.org/raster/sdm/6_sdm_methods.html#maxent)

```{r}
# downloaded http://www.cs.princeton.edu/~schapire/maxent
#   and placed into system.file("java", package="dismo")
# maxent() # MaxEnt version 3.4.3 

# *_lnd
cat   <- "landscape"
resp  <- "g_pres"
preds <- mdls_lnd$pred

dir_mx <- glue("data/maxent/{resp}_{cat}")
dir_create(dir_mx)

d_rp <- d %>% 
  select(all_of(resp), all_of(preds))

set.seed(42)
i_k     <- kfold(d_rp, 5)
d_train <- d_rp[i_k != 1, ]
d_test  <- d_rp[i_k == 1, ]

table(d_train[[resp]])
table(d_test[[resp]])

shelf(mrmaxent/maxnet)

v_resp    <- d_train %>% pull(resp)
#data = d_train %>% select(-all_of(resp), -asp_cir)
d_preds   <- d_train %>% select(-all_of(resp))
preds_fac <- d_preds %>% select_if(is.factor) %>% names()
d_preds   <- d_train %>% select(-all_of(resp)) %>% 
  mutate(
    across(where(is.factor), ~ as.integer(.x)))
# d_preds   <- d_preds %>% select(-all_of(preds_fac))

# levels(d_preds$asp_cir)
# as.integer(d_preds$asp_cir)

mx <- maxent(
  x       = d_preds, 
  p       = v_resp,
  factors = preds_fac,
  path    = dir_mx)
plot(mx)

response(mx)

e <- evaluate(
  p = d_test %>% filter(across(all_of(resp), ~ .x == 1)), 
  a = d_test %>% filter(across(all_of(resp), ~ .x == 0)), 
  mx)
e
threshold(e)

plot(e, 'ROC')
plot(e, 'TPR')
boxplot(e)
density(e)
```

### Random Forest

```{r}
library()

cat   <- "landscape"
resp  <- "g_pres"
preds <- mdls_lnd$pred

dir_mx <- glue("data/randomForest/{resp}_{cat}")
dir_create(dir_mx)

rf_train <- d_train %>% 
    mutate(g_pres = as.factor(g_pres))
rf_formula <- as.formula(glue("{resp} ~ ."))
rf <- randomForest(
  rf_formula,
  data = rf_train,
  importance=TRUE, proximity=T)

print(rf)

MDSplot(rf, rf_train[[resp]])

## Look at variable importance:
round(importance(rf), 2)

e <- evaluate(
  p = d_test %>% filter(across(all_of(resp), ~ .x == 1)), 
  a = d_test %>% filter(across(all_of(resp), ~ .x == 0)),
  rf)
e
```

```{r}
# ?randomForest
## Classification:
##data(iris)
set.seed(71)
iris.rf <- randomForest(Species ~ ., data=iris, importance=TRUE,
                        proximity=TRUE)
print(iris.rf)
## Look at variable importance:
round(importance(iris.rf), 2)
## Do MDS on 1 - proximity:
iris.mds <- cmdscale(1 - iris.rf$proximity, eig=TRUE)
op <- par(pty="s")
pairs(cbind(iris[,1:4], iris.mds$points), cex=0.6, gap=0,
      col=c("red", "green", "blue")[as.numeric(iris$Species)],
      main="Iris Data: Predictors and MDS of Proximity Based on RandomForest")
par(op)
print(iris.mds$GOF)

## The `unsupervised' case:
set.seed(17)
iris.urf <- randomForest(iris[, -5])
MDSplot(iris.urf, iris$Species)

## stratified sampling: draw 20, 30, and 20 of the species to grow each tree.
(iris.rf2 <- randomForest(iris[1:4], iris$Species, 
                          sampsize=c(20, 30, 20)))

# ?MDSplot

set.seed(1)
data(iris)
iris.rf <- randomForest(
  Species ~ ., iris,
  proximity=TRUE)
  # keep.forest=FALSE)
  #keep.forest=T)
MDSplot(iris.rf, iris$Species)

```


```{r}
# maxent
fnames <- list.files(path=paste(system.file(package="dismo"), '/ex', sep=''), 
              pattern='grd', full.names=TRUE )
predictors <- stack(fnames)
#plot(predictors)

# file with presence points
occurence <- paste(system.file(package="dismo"), '/ex/bradypus.csv', sep='')
occ <- read.table(occurence, header=TRUE, sep=',')[,-1]

# witholding a 20% sample for testing 
fold <- kfold(occ, k=5)
occtest <- occ[fold == 1, ]
occtrain <- occ[fold != 1, ]

# fit model, biome is a categorical variable
me <- maxent(predictors, occtrain, factors='biome')

# see the maxent results in a browser:
me

# use "args"
# me2 <- maxent(predictors, occtrain, factors='biome', args=c("-J", "-P"))

# plot showing importance of each variable
plot(me)


response(me)
```

TODO: consider cross-validation with [ENMeval Vignette](https://cran.r-project.org/web/packages/ENMeval/vignettes/ENMeval-vignette.html)



### BRT

```{r}
brt <- gbm.step(
  data=d_train, 
  gbm.x = 2:ncol(d_train), gbm.y = 1,
  family = "poisson"), 
  tree.complexity = 3,
  learning.rate = 0.01, bag.fraction = 0.5)
```

# Loop through models by response, category

$response \sim predictors_{category}$

```{r}
# moved to setup.R:
#   cols_cat  <- unique(d_cols$cat) %>% setdiff(c("_other", "_response")) %>% sort()
#   cols_resp <- cols_resp %>% setdiff("g_perf") %>% sort()
#   cols_pred
# TODO: show 2-col tbl of resp, is_binary "~" 2-col tbl of cat, pred "x" method: maxent (binary), RandomForest

# TODO: differentiate binary vs continuous, possibly showing table() or hist()
```

```{r}
set.seed(42)

for (cat in cols_cat){ # cat = "landscape"
  for (resp in cols_resp){ # resp = "g_pres"

    message(glue("cat:{cat}, resp:{resp}"))
    
    # get predictors based on significant, non-correlated predictors
    d_preds <- get_preds_sign_notcor(cat, resp)
    
    message(glue("{resp} ~ {paste(d_preds$pred, collapse =' + ')}"))
    
    if (nrow(d_preds) == 0)
      next
      
    # if response is binary, run maxent
    if (is_binary(d[[resp]])){
      message(glue("{cat}.{resp}.mx"))
      mx <- run_maxent(cat, resp, preds = d_preds$pred, use_cache = F)
    }
    
    # run randomForest
    message(glue("{cat}.{resp}.rf"))
    rf <- run_randomForest(cat, resp, preds = d_preds$pred, use_cache = F)
    
  } # end (resp in cols_resp)
} # end (cat in cols_cat)
```

# evaluate model performance per category

```{r}
resp_binary <- d %>% 
  select(all_of(cols_resp)) %>% 
  sapply(is_binary) %>% 
  which() %>% 
  names()
resp_cont <- setdiff(cols_resp, resp_binary)

suppressWarnings(rm(d_mdls))

mdls_cont <- expand.grid(
  cat  = cols_cat,
  resp = resp_cont) %>% 
  as_tibble() %>% 
  mutate(
    type    = "continous",
    path_rf = glue("{dir_models}/{cat}.{resp}.rf")) %>% 
  arrange(type, cat, resp) %>% 
  filter(dir_exists(path_rf)) %>% 
  select(type, cat, resp) %>% 
  rowwise() %>% 
  mutate(
    rf_NRMSE_sd     = get_nrmse(cat, resp, mdl="rf", method = "sd"), # other methods: iq, maxmin, mean
    predvars        = get_terms(cat, resp, mdl="rf")) %>%
  arrange(cat, rf_NRMSE_sd)

mdls_binary <- expand.grid(
  cat  = cols_cat,
  resp = resp_binary) %>% 
  as_tibble() %>% 
  mutate(
    type    = "binary",
    path_rf = glue("{dir_models}/{cat}.{resp}.rf"),
    path_mx = glue("{dir_models}/{cat}.{resp}.mx")) %>% 
  arrange(type, cat, resp) %>% 
  rowwise() %>% 
  mutate(
    rf_Accuracy = ifelse(
      dir_exists(path_rf),
      get_accuracy(cat, resp, mdl="rf"),
      NA),
    mx_Accuracy = ifelse(
      dir_exists(path_mx),
      get_accuracy(cat, resp, mdl="mx")),
    predvars        = get_terms(cat, resp, mdl="rf")) %>%
  select(type, cat, resp, rf_Accuracy, mx_Accuracy, predvars) %>% 
  arrange(type, cat, resp)

mdls <- bind_rows(
  mdls_cont,
  mdls_binary)

write_csv(mdls, here("data/models.csv"))
```


## Show outputs from single model

### maxent

```{r}
# params
resp = "g_pres"
cat = "landscape"

# get predictors and model outputs
d_preds <- get_preds_sign_notcor(cat, resp)
mx      <- run_maxent(cat, resp, d_preds$pred)

# summarize presence/absence in train/test data
table(mx$d_train[[resp]])
table(mx$d_test[[resp]])

# plot fit: variable contribution
plot(mx$maxent)

# plot fit: term plots
response(mx$maxent)

# plot evaluations
threshold(mx$evaluate)
plot(mx$evaluate, 'ROC')
plot(mx$evaluate, 'TPR')
boxplot(mx$evaluate)
density(mx$evaluate)
```

### randomForest, binary

```{r}
# params
resp = "g_pres"
cat = "landscape"

# get predictors and model outputs
d_preds <- get_preds_sign_notcor(cat, resp)
rf      <- run_randomForest(cat, resp, d_preds$pred)

print(rf$randomForest)

MDSplot(rf$randomForest, factor(rf$d_train[[resp]]))

## Look at variable importance:
round(importance(rf$randomForest), 2)

varImpPlot(rf$randomForest)

summary(rf)

rf$evaluate
```


### randomForest, continuous

```{r}
# params
resp = "g_cnt"
cat = "landscape"

# get predictors and model outputs
d_preds <- get_preds_sign_notcor(cat, resp)
rf      <- run_randomForest(cat, resp, d_preds$pred)

print(rf$randomForest)

# not relevant for continuous:
#  MDSplot(rf$randomForest, factor(rf$d_train[[resp]]))

## Look at variable importance:
round(importance(rf$randomForest), 2)
# TODO: convert to tibble w/ variable rowname and arrange(`%IncMSE`)

varImpPlot(rf$randomForest)

rf$evaluate

# show predictions for test & train
get_mdl_y(rf)
```


# Table of predictors

So far just for numeric predictors in the Plots worksheet.

```{r}
# d <- plot %>% 
#   filter(!is.na(gvtp_count)) %>% 
#   mutate(
#     gvtp = as.integer(!gvtp_count == "0")) %>% 
#   # TODO: handle non-numeric
#   select_if(is.numeric) %>% 
#   mutate(
#     gvtp = gvtp == 1)

#table(d$gvtp)



#predictors <- setdiff(names(d), c("plot", "gvtp")) # %>% sort()

# d_long <- d %>% 
#   pivot_longer(-gvtp, "variable") # %>%
  # filter(variable %in% (unique(.$variable) %>% head(5))) # %>% 
  # mutate(
  #   gvtp = as.factor(gvtp)) %>% 
  #arrange(variable, gvtp, value)

# d_long$var <- abbreviate(d_long$variable, minlength=4)

# d_avg <- d_long %>% 
#   select(-variable) %>% 
#   group_by(var, gvtp) %>% 
#   summarize(
#     min     = min(value, na.rm=T), 
#     avg     = mean(value, na.rm=T), 
#     max     = max(value, na.rm=T),
#     sd      = sd(value, na.rm=T),
#     .groups="drop")

# d_smry <- d_avg %>% 
#   pivot_wider(
#     -sd, 
#     names_from = gvtp, 
#     names_glue = "{ifelse(gvtp, 'p', 'a')}_{.value}",
#     values_from = c(min, avg, max))

#d_pres_num_lng

var_smry <- d_pres_num_avg %>% 
  pivot_wider(
    -sd, 
    names_from = g_pres, 
    names_glue = "{ifelse(g_pres==1, 'p', 'a')}_{.value}",
    values_from = c(min, avg, max))

plot_var <- function(var, redo = F){
  # var = "cu"
  v_smry <- var_smry %>% 
    filter(var == !!var) 
  
  img <- here(glue("vars_imgs/{var}.png"))
  
  if (file.exists(img) & !redo)
    return(img)
  
  png(img, width = 4.8, height = 1.2, units="in", res=72)
  p <- d_pres_num_lng %>% 
    filter(var == !!var) %>% 
    mutate(
       g_pres = factor(c("absent", "present")[g_pres+1])) %>% 
    ggplot(aes(x=value, fill=g_pres, color=g_pres)) +
    geom_density(alpha=0.2) +
    # geom_histogram(
    #   aes(y=..density..), position="identity", alpha=0.5) +
    geom_vline(
      data = d_pres_num_avg %>% 
        mutate(
          g_pres = factor(c("absent", "present")[g_pres+1])) %>%
        filter(var == !!var),
      aes(xintercept=avg, color=g_pres),
      linetype="solid", lwd=2) +
    theme_nothing() + 
    scale_x_continuous(expand=c(0,0)) +
    scale_y_continuous(expand=c(0,0)) +
    labs(x = NULL, y = NULL)
  print(p)
  #ggsave(png, width=2, height=1, units="in", dpi=72)
  dev.off()
  
  img
}
```


```{r}
var_smry <- var_smry %>% 
  mutate(
    nchar   = nchar(var),
    Variable = strtrim(var, 26),
    Absence  = glue("{signif(a_avg, 3)} ({signif(a_min, 3)} - {signif(a_max, 3)})"),
    Presence = glue("{signif(p_avg, 3)} ({signif(p_min, 3)} - {signif(p_max, 3)})"),
    Histogram = "",
    png = map_chr(var, plot_var, redo = T))

ft <- flextable(
  var_smry, 
  col_keys = c("Variable", "Absence", "Presence", "Histogram"))
ft <- compose(
  ft, i = 1:nrow(var_smry), j = "Histogram", 
  value = as_paragraph(as_image(src = png, width = 1.2, height = 0.3)))
ft <- autofit(ft)
ft
```


